Casting doubt on the causal link between intelligence and age at first intercourse:\\ A cross-generational sibling comparison design using the NLSY

### 2016-06-17 Manuscript/Manuscript.tex
fixed tables
resubmitted

\documentclass[a4paper,man,apacite,natbib,12pt,longtable]{apa6}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage{../Common/style/Mason}
\usepackage{listings}
\usepackage{inconsolata}

%%%%%%%%%%%% Title %%%%%%%%%%%%
\title{Is the Link Between Intelligence and Age at First Intercourse Causal or Correlational? \\ A Cross-Generational Sibling Comparison Design Using the NLSY}
\shorttitle{Intelligence and AFI}
%
% Authors and Affiliations
\author{S. Mason Garrison and Joseph Lee Rodgers}
\affiliation{Vanderbilt University}
%
% Author Note
\authornote{\input{../Common/content/title/authornote.tex}}
%
% Abstract
\abstract{Last compiled \today\ at \currenttime\\
\input{../Common/content/title/abstract_empreport.txt}}
%
% Key Words
\keywords{Age at first intercourse; Cross-sectional Data; Discordant Sibling Design; Intelligence; Quasi-Experimental; Siblings}
%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle
%
%%%%%%%%%%%% R %%%%%%%%%%%%

@@

%%%%%%%%%%%% Method %%%%%%%%%%%%
\section{Method}
\subsection{Research Design}
\input{../Common/content/writing/design.tex}
%
% Sample
\subsection{Sample}
\input{../Common/content/writing/subjects.tex}
%
% Tetrads
\subsection{Tetrads}
\input{../Common/content/writing/tetrad.tex}
%
%%%%%%%%%%%% Measurement %%%%%%%%%%%%
% Age at First Intercourse
\subsection{Measures}
\input{../Common/content/writing/AFI.tex}
\noindent\begin{minipage}{\linewidth}
\captionof{figure}{Smoothed Density Plots of AFI by Race and Sex}
\label{plot_afi_by_race_sex}
\begin{center}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.8\paperwidth]{figure/plot_afi_by_race_sex-1}

@@

Our design assumes that the difference scores of our measures are reliable. We've reported the test-retest reliability of Gen2 intelligence and Gen1 AFI in earlier sections. Here, we report the test-retest reliability of the pairwise differences of those measures. 
\subsubsection{AFI} Comparing sibling differences in AFI as reported in 1983 and 1984 (n = 783 pairs) we found a moderate correlation (r = .76). The sample of sibling pairs with complete information in 1985 was too small (n = 12 pairs) to compare to the other two years. Regardless, sibling differences in self-reported AFI appear satisfactory for research purposes. We note that the common concern with unreliability of difference scores is at least partially mitigated when the separate scores defining the differences are reliable themselves. Because we could not estimate test-retest reliabilities for Generation 2, we calculated the reliability using Lord's equation (\citeyear{Lord1963}). Generation 2 Mean AFI difference scores were also moderately reliable (r = 0.734) and comparable to Generation 1 sibling differences.

\subsubsection{Intelligence} Cousin differences in intelligence as assessed at ages 9.5, 10.5, and 11.5 were correlated using three different linking methods (Mixed, Daughters, Sons). Reliabilities across linking methods was consistent and high (min r = .86; max r = .95). However, we could not estimate test-retest reliabilities for Generation 1; we calculated the reliability using Lord's equation (\citeyear{Lord1963}). The calculated reliability of Generation 1's differences in AFQT was .61, which was only minimally  acceptable, and obviously lower than the empirical correlation we derived for cousin differences.%0.606

%%%%%%%%%%%% Results %%%%%%%%%%%%
\section{Results}
\input{../Common/content/writing/results.tex}
%
% Between-Family
\subsection{Between-Family Analyses}
\input{../Common/content/writing/results_between.tex}
%
% Within-Family
\subsection{Within-Family Analyses}
\input{../Common/content/writing/results_within.tex}
%
%%%%%%%%%%%% Discussion %%%%%%%%%%%%
\section{Discussion}
\input{../Common/content/writing/discussion.tex}

%%%%%%%%%%%% References %%%%%%%%%%%%
\bibliography{Projects-AFI}
  %%%%%%%%%%%% Appendix %%%%%%%%%%%%
   \begin{landscape} \appendix\label{appen}
  %
  % Age 10.5 Replication
  \section{Age 10.5 Replication}\label{appen10}
  %% Between-Family

@@

  \caption{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Mom_Intelligence_Mean_Child_AFI_10}

@@

  \hline \\[-1.8ex] 

@@

  \caption{Gen2 Mean Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Child_Intelligence_Mean_Child_AFI_10}

@@

  \caption{Mean Joint Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Joint_Intelligence_Mean_Child_AFI_10}

@@

  \caption{Gen1 Dif Intelligence $\rightarrow$ Gen2 Dif AFI}\label{table_Dif_Mom_Intelligence_Dif_Child_AFI_10}

@@

  \hline \\[-1.8ex] 
  & \multicolumn{3}{c}{\textit{Dependent variable:} Difference in Gen2 Mean AFI} \\ 

@@

  \caption{Gen2 Dif Intelligence $\rightarrow$ Gen2 Dif AFI}\label{table_Dif_Child_Intelligence_Dif_Child_AFI_10}

@@

  \caption{Joint Dif Intelligence $\rightarrow$ Gen2 Dif AFI}\label{table_Dif_Joint_Intelligence_Dif_Child_AFI_10}

@@

  \caption{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Mom_Intelligence_Mean_Child_AFI_11}

@@

  \caption{Gen2 Mean Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Child_Intelligence_Mean_Child_AFI_11}

@@

  \caption{Mean Joint Intelligence $\rightarrow$ Gen2 Mean AFI}\label{table_Mean_Joint_Intelligence_Mean_Child_AFI_11}

@@

  \caption{Gen1 Dif Intelligence $\rightarrow$ Gen2 Dif AFI}\label{table_Dif_Mom_Intelligence_Dif_Child_AFI_11}

@@

  \hline \\[-1.8ex] 
  & \multicolumn{3}{c}{\textit{Dependent variable:} Differences in Gen2 Mean AFI} \\ 

@@

  \caption{Gen2 Dif Intelligence $\rightarrow$ Gen2 Dif AFI}\label{table_Dif_Child_Intelligence_Dif_Child_AFI_11}

@@

  \caption{Dif Joint Intelligence $\rightarrow$ Gen2 Dif AFI}\label{table_Dif_Joint_Intelligence_Dif_Child_AFI_11}

@@

  %

>>>>

\documentclass[a4paper,man,apacite,natbib,12pt,longtable,mask]{apa6}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage{../Common/style/Mason}
\usepackage{listings}
\usepackage{inconsolata}
\usepackage[bottom]{footmisc} %fix footnotes
\usepackage{lineno}
\usepackage{adjustbox}
%\linenumbers
\interfootnotelinepenalty=10000
%%%%%%%%%%%% Title %%%%%%%%%%%%
\title{Casting doubt on the causal link between intelligence and age at first intercourse:\\ A cross-generational sibling comparison design using the NLSY}
\shorttitle{Intelligence and AFI}
%
% Authors and Affiliations
\author{S. Mason Garrison and Joseph Lee Rodgers}
\affiliation{Vanderbilt University}
%
% Author Note
\authornote{\input{../Common/content/title/authornote.tex}}
%
% Abstract
\abstract{%Last compiled \today\ at \currenttime\\
\input{../Common/content/title/abstract_empreport.txt}}
%
% Key Words
\keywords{Age at first intercourse; Cross-sectional Data; Discordant Sibling Design; Intelligence; Quasi-Experimental; Siblings}
%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle

%
%%%%%%%%%%%% R %%%%%%%%%%%%

@@

%%%%%%%%%%%% Method %%%%%%%%%%%%
\section{Method}
\subsection{Research Design and Analytic Approach}
\input{../Common/content/writing/design.tex}
%
% Sample
\subsection{Sample}
\input{../Common/content/writing/subjects.tex}
%
% Tetrads
\subsection{Tetrads}
\input{../Common/content/writing/tetrad.tex}
%
%%%%%%%%%%%% Measurement %%%%%%%%%%%%
% Age at First Intercourse
\subsection{Measures}
\input{../Common/content/writing/AFI.tex}
\noindent\begin{minipage}{\linewidth}
\captionof{figure}{Smoothed Density Plots of Gen1 AFI by Race and Sex}
\label{plot_afi_by_race_sex}
\begin{center}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.65\paperwidth]{figure/plot_afi_by_race_sex-1}

@@

Our design assumes that the difference scores of our measures are reliable. We have reported the test-retest reliability of Gen2 intelligence and Gen1 AFI in earlier sections. Here, we report the test-retest reliability of the pairwise differences of those measures. 
\subsubsection{AFI} Comparing sibling differences in AFI as reported in 1983 and 1984 (n = 783 pairs) we found a moderate correlation (r = 0.76). The sample of sibling pairs with complete information in 1985 was too small (n = 12 pairs) to compare to the other two years. Regardless, sibling differences in self-reported AFI appear satisfactory for research purposes. We note that the common concern with unreliability of difference scores is at least partially mitigated when the separate scores defining the differences are reliable themselves. Because we could not estimate test-retest reliabilities for Generation 2, we calculated the reliability using Lord's (\citeyear{Lord1963}) equation. Generation 2 Mean AFI difference scores were also moderately reliable (r = 0.73) and comparable to Generation 1 sibling differences.

\subsubsection{Intelligence} Cousin differences in intelligence as assessed at ages 9.5, 10.5, and 11.5 were correlated using three different linking methods (Mixed, Daughters, Sons). Reliabilities across linking methods was consistent and high (min r = .86; max r = .95). However, we could not estimate test-retest reliabilities for Generation 1; we calculated the reliability using Lord's (\citeyear{Lord1963}) equation. The calculated reliability of Generation 1's differences in AFQT was 0.70, which was marginally acceptable, and obviously lower than the empirical correlation we derived for cousin differences.%0.7

%%%%%%%%%%%% Results %%%%%%%%%%%%
\section{Results}
\input{../Common/content/writing/results.tex}
%
% Between-Family
\subsection{Between-Family Analyses}
\input{../Common/content/writing/results_between.tex}

\noindent\begin{minipage}{\paperwidth}
\captionof{figure}{Gen2 AFI vs. Gen1 \& Gen2 Intelligence}
\label{plot_gen2_afi}
\begin{center}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=1\paperwidth]{figure/plot_gen2_afi-1}

\end{knitrout}
\end{center}
\end{minipage}
\end{landscape}

%
% Within-Family
\subsection{Combination Between- and Within-Family Analyses}
\input{../Common/content/writing/results_within.tex}
%
\begin{landscape}
\noindent\begin{minipage}{\linewidth}
\captionof{figure}{Gen1 AFI vs. Gen1 Intelligence with Marginal Distributions by Sibling Difference in AFQT}
\label{plot_gen1_afi}
\begin{center}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=.8\paperwidth]{figure/plot_gen1_afi-1}

\end{knitrout}
\end{center}\end{minipage}\end{landscape}

%%%%%%%%%%%% Discussion %%%%%%%%%%%%
\section{Discussion}
\input{../Common/content/writing/discussion.tex}
%%%%%%%%%%%% References %%%%%%%%%%%%
\bibliography{Projects-AFI}
%%%%%%%%%%%% Appendix %%%%%%%%%%%%
   \begin{landscape} \appendix\label{appen}
  %
  % Age 10.5 Replication
  \section{Age 10.5 Replication}\label{appen10}
  %% Between-Family

@@

  \caption{Between-Family: Gen1 Intelligence Predicts Gen2 AFI} \label{table_Mean_Mom_Intelligence_Mean_Child_AFI_10}

@@

  \hline \\[-3.8ex] 

@@

  \caption{Between-Family: Gen2 Intelligence Predicts Gen2 AFI} \label{table_Mean_Child_Intelligence_Mean_Child_AFI_10}

@@

  \caption{Between-Family: Gen1 \& Gen2 Intelligence Predict Gen2 AFI} \label{table_Mean_Joint_Intelligence_Mean_Child_AFI_10}

@@

  \caption{Within-Family: Gen1 Differences in Intelligence Predict Gen2 Differences in AFI} \label{table_Dif_Mom_Intelligence_Dif_Child_AFI_10}

@@

  \hline \\[-3.8ex] 
  & \multicolumn{3}{c}{\textit{Dependent variable:} Difference in Gen2 AFI} \\ 

@@

  \caption{Within-Family: Gen2 Differences in Intelligence Predict Gen2 Differences in AFI} \label{table_Dif_Child_Intelligence_Dif_Child_AFI_10}

@@

  \caption{Within-Family: Gen1 \& Gen2 Differences in Intelligence Predict Gen2 Differences in AFI} \label{table_Dif_Joint_Intelligence_Dif_Child_AFI_10}

@@

  \caption{Between-Family: Gen1 Intelligence Predicts Gen2 AFI} \label{table_Mean_Mom_Intelligence_Mean_Child_AFI_11}

@@

  \caption{Between-Family: Gen2 Intelligence Predicts Gen2 AFI} \label{table_Mean_Child_Intelligence_Mean_Child_AFI_11}

@@

  \caption{Between-Family: Gen1 \& Gen2 Intelligence Predict Gen2 AFI} \label{table_Mean_Joint_Intelligence_Mean_Child_AFI_11}

@@

  \caption{Within-Family: Gen1 Differences in Intelligence Predict Gen2 Differences in AFI}\label{table_Dif_Mom_Intelligence_Dif_Child_AFI_11}

@@

  \hline \\[-3.8ex] 
  & \multicolumn{3}{c}{\textit{Dependent variable:} Differences in Gen2 AFI} \\ 

@@

  \caption{Within-Family: Gen2 Differences in Intelligence Predict Gen2 Differences in AFI} \label{table_Dif_Child_Intelligence_Dif_Child_AFI_11}

@@

  \caption{Within-Family: Gen1 \& Gen2 Differences in Intelligence Predict Gen2 Differences in AFI} \label{table_Dif_Joint_Intelligence_Dif_Child_AFI_11}

@@

### 2016-06-17 Manuscript/highlights.Rnw
resubmitted

\begin{document}
\shorttitle{Intelligence and AFI}
\begin{itemize}
\item The causal relationship between intelligence and age at first intercourse is tested, using inter-generational sibling comparisons.
\item When we evaluated the relationship between maternal/child intelligence and AFI, using a between-family design, we found a positive relationship between intelligence and AFI.
\item When we use a within-family design and compare offspring of sisters, and then compare the offspring themselves, the relationship between maternal/child intelligence and AFI virtually disappears.
\item These results suggest that the cause of the intelligence-AFI link is not intelligence per se, but rather differences between families (parental education, SES, \textit{etc.}) that correlate with family-level (but not individual-level) intelligence.
\end{itemize}
\end{document}

>>>>

\begin{document}
\shorttitle{Intelligence and AFI}
\begin{itemize}
\item We tested the causal link between AFI and intelligence using inter-generational data, and explicitly parsed between- and within-family variance.

\item When we evaluated the relationship using a between-family design, we found a positive relationship between intelligence and AFI for both maternal intelligence and child intelligence.

\item When we use a within-family design, the relationship between intelligence and AFI vanishes for both maternal intelligence and child intelligence.

\item We find no evidence for intelligence being a direct causal influence on AFI.

\item These results suggest that the cause of the intelligence-AFI link is not intelligence \textit{per se}, but rather differences between families (parental education, SES, \textit{etc.}) that correlate with family-level (but not individual-level) intelligence.

\item Although ``Smart Teens Don't Have Sex (or Kiss Much Either)'' (Halpern \textit{et al}, 2000) at a descriptive level, the specific and proximal causal factors do not appear to be much related to smartness.

\end{itemize}
\end{document}

### 2016-06-17 Manuscript/highlights.tex
resubmitted

\usepackage{alltt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\shorttitle{Intelligence and AFI}
\begin{itemize}
\item The causal relationship between intelligence and age at first intercourse is tested, using intergenerational sibling comparisons.
\item When we evaluated the relationship between maternal/child intelligence and AFI, using a between-family design, we found a positive relationship between intelligence and AFI.
\item When we use a within-family design and compare offspring of sisters, and then compare the offspring themselves, the relationship between maternal/child intelligence and AFI virtually disappears.
\item These results suggest that the cause of the intelligence-AFI link is not intelligence per se, but rather differences between families (parental education, SES, \textit{etc.}) that correlate with family-level (but not individual-level) intelligence.
\end{itemize}
\end{document}

>>>>

\usepackage{alltt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\shorttitle{Intelligence and AFI}
\begin{itemize}
\item We tested the causal link between AFI and intelligence using inter-generational data, and explicitly parsed between- and within-family variance.

\item When we evaluated the relationship using a between-family design, we found a positive relationship between intelligence and AFI for both maternal intelligence and child intelligence.

\item When we use a within-family design, the relationship between intelligence and AFI vanishes for both maternal intelligence and child intelligence.

\item We find no evidence for intelligence being a direct causal influence on AFI.

\item These results suggest that the cause of the intelligence-AFI link is not intelligence \textit{per se}, but rather differences between families (parental education, SES, \textit{etc.}) that correlate with family-level (but not individual-level) intelligence.

\item Although ``Smart Teens Don't Have Sex (or Kiss Much Either)'' (Halpern \textit{et al}, 2000) at a descriptive level, the specific and proximal causal factors do not appear to be much related to smartness.

\end{itemize}
\end{document}

### 2016-06-17 Manuscript/title.Rnw
resubmitted

%%%%%%%%%%%% Title %%%%%%%%%%%%
\title{Is the Link Between Intelligence and Age at First Intercourse Causal or Correlational? \\ A Cross-Generational Sibling Comparison Design Using the NLSY}
\shorttitle{Intelligence and AFI}
%
% Authors and Affiliations
\author{S. Mason Garrison and Joseph Lee Rodgers}
\affiliation{Vanderbilt University}
%
% Author Note
\authornote{\input{../Common/content/title/authornote.tex}}
%
% Abstract
\abstract{%Last compiled \today\ at \currenttime\\
\input{../Common/content/title/abstract_empreport.txt}}
%
% Key Words
\keywords{Age at first intercourse; Cross-sectional Data; Discordant Sibling Design; Intelligence; Quasi-Experimental; Siblings}
%
\begin{document}
\maketitle
\end{document}

>>>>

%%%%%%%%%%%% Title %%%%%%%%%%%%
\title{Casting doubt on the causal link between intelligence and age at first intercourse:\\ A cross-generational sibling comparison design using the NLSY}
\shorttitle{Intelligence and AFI}
%
% Authors and Affiliations
\author{S. Mason Garrison and Joseph Lee Rodgers}
\affiliation{Vanderbilt University}
%
% Author Note
\authornote{\input{../Common/content/title/authornote.tex}}
%
% Abstract
\abstract{%Last compiled \today\ at \currenttime\\
\input{../Common/content/title/abstract_empreport.txt}}
%
% Key Words
\keywords{Age at first intercourse; Cross-sectional Data; Discordant Sibling Design; Intelligence; Quasi-Experimental; Siblings}
%
\begin{document}
\maketitle
\end{document}
